The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables with \pkg{vcd}

### 2005-10-25 trunk/vcd/paper/JSS.Rnw
revision including AZ fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@485 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% almost as usual
\author{David Meyer\\Wirtschaftsuniversit\"at Wien \And 
        Achim Zeileis\\Wirtschaftsuniversit\"at Wien \And 
        Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\title{The Strucplot Framework:\\ Visualizing Multi-way Contingency
  Tables With \pkg{vcd}}

@@

%% an abstract and keywords
\Abstract{
This paper describes the `strucplot' framework for the visualization of
multi-way contingency tables. Strucplot displays include hierarchical
conditional plots such as mosaic, association, and sieve plots, and
can be combined into more complex, specialized plots for visualizing
conditional independence, GLMs, and the results of
independence tests. The frameworks' modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of graphical
appearance control (`grapcon') functions. The framework is provided by
the \pkg{vcd} package freely available from the
Comprehensive \proglang{R} Archive Network (CRAN).
}

\Keywords{contingency tables, mosaic plots, association plots, sieve plots, \proglang{R},
categorical data, independence, conditional independence, HSV, HCL,
residual-based shading, \pkg{grid}}
\Plainkeywords{contingency tables, mosaic plots, association plots,
  sieve plots, R, categorical data, independence, 
  conditional independence, HSV, HCL, residual-based shading, grid}

@@

  Achim Zeileis\\
  Department of Statistics \& Mathematics\\
  E-mail: \email{Achim.Zeileis@wu-wien.ac.at}\\
  URL: \url{http://www.ci.tuwien.ac.at/~zeileis/}\\

  Kurt Hornik\\
  Department of Statistics \& Mathematics\\
  E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
  URL: \url{http://www.wu-wien.ac.at/cstat/hornik/}\\

@@

In order to explain multi-dimensional categorical data, statisticians
typically look for (conditional) independence structures. Whether the
task is purely exploratory or model based, techniques such as mosaic and
association plots offer good support for visualization. Both have been
extended over the last two decades, and implementations exist in many
statistical environments. 
Both graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \citep{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. An \emph{association plot} \citep{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

@@

\begin{enumerate}
\item Varying the shape of bar plots and mosaic displays to yield, e.g.,
  double-decker plots \citep{vcd:hofmann:2001}, spine plots, or spinograms.
\item Using residual-based shadings to visualize log-linear models
  \citep{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \citep{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \citep{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \citep{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \citep{vcd:Meyer+Zeileis+Hornik:2003}.
\end{enumerate}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS}, \proglang{ViSta}, \proglang{MANET}, \proglang{Mondrian},
\proglang{R}, and \proglang{S-PLUS}. Table \ref{tab:compare} gives an overview of the
available functionality in these systems.

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \proglang{ViSta} & \proglang{MANET}/\proglang{Mondrian}\\\hline
    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
    Shape                    &     &        & $\times$ &       & $\times$\\
    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
    Interaction              &     &        &   &   $\times$   & $\times$\\
    Extensible Design        &     &        & $\times$ &       &  \\\hline
  \end{tabular}
  \caption{Comparison of current software environments}
  \label{tab:compare}
\end{table}

Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
extensions. Figure \ref{fig:arthritis} shows the results from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, using an
extended mosaic plot with residual-based shading based on the
maximum statistic: clearly, the new treatment is effective. Figure
\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
by the means of a conditional association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null) causes the rejection of the
hypothesis of conditional independence. Figure \ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status. The Chi-squared
test of independence rejects the null hypothesis: possibly, because
too many men who had premarital sex also had extramarital sex?
Finally, Figure \ref{fig:titanic} visualizes the `Survival on the Titanic' 
data using a double-decker plot. Here, a binary response (survival to the disaster)
shall be explained by other factors (class, gender, and age). The blue
boxes represent the proportion of survived passengers in a particular
stratum. The proportions of saved women and children are indeed higher
than those of men, but they clearly decrease from the 1st to the 3rd class. In
addition, the proportion of saved men in the 1st class is higher than
in the others.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(UCBAdmissions, panel = cotab_assoc)
@
\caption{Conditional association plot for the \data{UCBAdmissions} data.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

@@

This paper describes the strucplot framework provided by the \pkg{vcd} package
for the \proglang{R} system and environment for statistical computing
and graphics, available from the Comprehensive \proglang{R} Archive Network (CRAN).
The framework integrates displays such as mosaic, association, and
sieve plots by their common property of being 
flat representations of contingency tables.
These basic plots, as well as specialized displays for conditional
independence, can be used both for exploratory visualization and
model-based analysis. Exploratory techniques include specialized displays for
the bivariate case, as well as pairs and trellis plot-like displays for
higher-dimensional tables. Model-based tools include 
methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (logistic regression,
log-linear models). Additionally, two of the frameworks' further strengths are its
flexibility and extensibility: graphical appearance aspects such as shading,
labeling, and spacing are modularized by means of `graphical
appearance functions', allowing fine-granular customization and
user-level extensions.

The reminder of the paper is organized as follows. In
Section \ref{sec:strucplot}, we give an overview of the strucplot
framework, describing the hierarchy of the main components and the
basic functionality. 
In Section \ref{sec:shading}, we demonstrate how
residual-based shadings support the visualization of log-linear models and
the results of independence tests. Also, we explain step-by-step how
the concepts of generating functions and graphical appearance control
functions can be combined to provide a flexible customization of
complex graphical displays as created by the strucplot framework.
Sections \ref{sec:labeling} and \ref{sec:spacing} discuss in detail the labeling
and spacing features, respectively. Section \ref{sec:conclusion}
finally concludes the work.

@@

The strucplot framework is highly modularized: Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot. These are examples of `graphical appearance control'
(`grapcon') functions. They are created by generating functions
(`grapcon generators'), allowing
flexible parameterization and extensibility (Figure \ref{fig:struc}
only shows the generators).
The first part of the generator names (\code{\emph{group\_foo}()})
reflects the group they belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\emph{foo}()}, 
\code{labeling\_\emph{foo}()}, and \code{legend\_\emph{foo}()}) 
directly produce graphical output (``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\emph{foo}()} and \code{shading\_\emph{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\emph{foo}()} implement the core functionality, creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulas. Finally, on the fourth
level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

@@

As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous variable (sex, i.e., gender). The
`flattened' contingency table can be obtained using the
\codefun{structable} function (quite similar to \codefun{ftable} in
base \proglang{R}, but allowing the specification of split directions):

@@

Let us first visualize the contingency table by means of a mosaic plot
\citep{vcd:Hartigan+Kleiner:1984} which is basically 
an area-proportional visualization of (typically, observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. Figure \ref{fig:observed} depicts 
the effect of

@@

%\setkeys{Gin}{width=0.75\textwidth}
\begin{figure}[p]
\begin{center}
<<Observedfig,fig=TRUE,echo=FALSE>>=
<<Observed>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data.}
\label{fig:observed}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<Expectedfig,fig=TRUE,echo=FALSE>>=
<<Expected>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}
%\setkeys{Gin}{width=0.7\textwidth}

@@

\noindent For compatibility with \codefun{mosaicplot} in base \proglang{R},
the \codefun{mosaic} function also allows the use of a \code{"direction"} 
argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figure \ref{fig:split}):

@@

So far, we have visualized full tables. For objects of class \class{table},
conditioning on levels (i.e., choosing a table subset for fixed levels
of the conditioning variable(s)) is simply done by indexing.
However, subsetting \code{"structable"} objects is more restrictive 
because of their inherent conditional structure. Since the variables
on both the row and the columns side are nested, 
conditioning is only possible ``outside-in'':

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=12>>=
<<Variables>>
@
\caption{Distribution of hair and eye color, given gender.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
<<Conditioning>>
@
\caption{Mosaic plot for conditional independence structures.}
\label{fig:conditioning}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
<<cotabplot>>
@
\caption{Conditional table plot for the \data{HairEyeColor} data.}
\label{fig:cotabplot}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

This flexible way of specifying graphical parameters is the basis for a suite of
shading functions that modify the tiles' appearance with respect to a vector of 
residuals, resulting from deviations of observed from expected
values under a given log-linear model. 
The idea is to visualize at least sign and absolute size of the residuals, but some
shadings, additionally, indicate overall significance. One particular
shading, the maximum shading, even allows to identify those cells that
cause the rejection of the null hypothesis.

@@

\begin{figure}[p]
\begin{center}
<<continuousfig,fig=TRUE,echo=FALSE>>=
<<continuous2>>
@
\caption{The \data{HairEyeColor} data with continuous shading.}
\label{fig:continuous}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<bundesligafig,fig=TRUE,echo=FALSE>>=
<<bundesliga>>
@
\caption{Non-significant $\chi^2$ test using part of the \data{Bundesliga} data.}
\label{fig:bundesliga}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
<<friendlyfig,fig=TRUE,echo=FALSE>>=
<<friendly>>
@
\caption{The \data{Bundesliga} data for 1995 using the Friendly
  shading and a legend with fixed bins.}
\label{fig:friendly}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<hclshading1fig,fig=TRUE,echo=FALSE>>=
<<hclshading1>>
@
\caption{The \data{HairEyeColor} data, using default HCL color palette.}
\label{fig:hclshading1}
\end{center}
\end{figure}

\begin{figure}[p]
\begin{center}
<<hclshading2fig,fig=TRUE,echo=FALSE>>=
<<hclshading2>>
@
\caption{The \data{HairEyeColor} data, using a custom HCL color palette.}
\label{fig:hclshading2}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<rotatefig,fig=TRUE,echo=FALSE>>=
<<rotate>>
@
\caption{Rotating labels.}
\label{fig:rotating}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<repeatfig,fig=TRUE,echo=FALSE>>=
<<repeat>>
@
\caption{Inhibiting the repetition of levels.}
\label{fig:repeat}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<marginsfig,fig=TRUE,echo=FALSE>>=
<<margins>>
@
\caption{Changes in the margins.}
\label{fig:margins}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<boxesfig,fig=TRUE,echo=FALSE>>=
<<boxes>>
@
\caption{Boxes and Clipping.}
\label{fig:boxes}
\end{center}
\end{figure}

@@

Spacing of strucplot displays is customizable in a similar way than shading. The
\code{spacing} argument of the \codefun{strucplot} function takes a
list of \class{unit} vectors, one for each dimension, specifying the
space between the tiles corresponding to the levels. As an example,
consider the introductory example of the \data{Arthritis} data 
(Figure \ref{fig:arthritis}). Since we are interested in the effect of
the medicament in the placebo and treatment groups, a mosaic plot is
certainly appropriate to visualize the three levels of `Improved' in
the two `Treatment' strata. However, following a slightly different approach,
we might as well want to \emph{highlight} patients with `Marked' improvement in both
groups, that is, produce a spine plot. To obtain such a display within 
the strucplot framework\footnote{The \pkg{vcd} package also provides 
the \codefun{spine} function for spine plots and spinograms.}, it
suffices to set the space between the `Improved' tiles to 0 (see
Figure \ref{fig:artspine}):

<<artspine,echo=TRUE,results=verbatim>>=
(art <- structable(~Treatment + Improved, data = Arthritis, split_vertical = TRUE))
(my_spacing <- list(unit(0.5, "lines"), unit(c(0, 0), "lines")))
my_colors <- gpar(fill = c("lightgray", "darkgray", "black"))
mosaic(art, spacing = my_spacing, gp = my_colors)
@

@@

\begin{appendix}
\section{Data sets}

>>>>

%% almost as usual
\author{David Meyer\\Wirtschaftsuniversit\"at Wien \And 
        Achim Zeileis\\Wirtschaftsuniversit\"at Wien \And 
        Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\title{The Strucplot Framework:\\ Visualizing Multi-way Contingency
  Tables with \pkg{vcd}}

@@

%% an abstract and keywords
\Abstract{
This paper describes the `strucplot' framework for the visualization of
multi-way contingency tables. Strucplot displays include hierarchical
conditional plots such as mosaic, association, and sieve plots, and
can be combined into more complex, specialized plots for visualizing
conditional independence, GLMs, and the results of
independence tests. The frameworks' modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of \underline{gr}aphical
\underline{ap}pearance \underline{con}trol (`grapcon') functions. 
The framework is provided by the \proglang{R} package \pkg{vcd}.
}

\Keywords{contingency tables, mosaic plots, association plots, sieve plots, 
categorical data, independence, conditional independence, HSV, HCL,
residual-based shading, \pkg{grid}, \proglang{R}}
\Plainkeywords{contingency tables, mosaic plots, association plots,
  sieve plots, categorical data, independence, 
  conditional independence, HSV, HCL, residual-based shading, grid, R}

@@

  Achim Zeileis\\
  Department of Statistics and Mathematics\\
  E-mail: \email{Achim.Zeileis@wu-wien.ac.at}\\
  URL: \url{http://www.ci.tuwien.ac.at/~zeileis/}\\

  Kurt Hornik\\
  Department of Statistics and Mathematics\\
  E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
  URL: \url{http://www.wu-wien.ac.at/cstat/hornik/}\\

@@

In order to explain multi-dimensional categorical data, statisticians
typically look for (conditional) independence structures. Whether the
task is purely exploratory or model-based, techniques such as mosaic and
association plots offer good support for visualization. Both have been
extended over the last two decades, and implementations exist in many
statistical environments. 
Both graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \citep{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically, observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. An \emph{association plot} \citep{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

@@

\begin{enumerate}
\item Varying the shape of bar plots and mosaic displays to yield, e.g.,
  double-decker plots \citep{vcd:hofmann:2001}, spine plots, or
  spinograms \citep{vcd:hofmann+theus}.
\item Using residual-based shadings to visualize log-linear models
  \citep{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \citep{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models 
  \citep{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \citep{vcd:Meyer+Zeileis+Hornik:2003}.
\end{enumerate}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \citep{vcd:SAS:2005}, \proglang{ViSta} \citep{vcd:young:1996}, 
\proglang{MANET} \citep{vcd:Unwin+Hawkins+Hofmann:1996},
\proglang{Mondrian} \citep{vcd:Theus:2003},
\proglang{R} \citep{vcd:R:2005}, and \proglang{S-PLUS} \citep{vcd:SPLUS:2005}. 
Table \ref{tab:compare} gives an overview of the available functionality in these systems.

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET}/\pkg{Mondrian}\\\hline
    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
    Shape                    &     &        & $\times$ &       & $\times$\\
    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
    Interaction              &     &        &   &   $\times$   & $\times$\\
    Extensible Design        &     &        & $\times$ &       &  \\\hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}

Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
extensions (the code is shown in Appendix \ref{sec:code}). 
Figure \ref{fig:arthritis} shows the results from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, using an
extended mosaic plot with residual-based shading based on the
maximum statistic: clearly, the new treatment is effective. Figure
\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
by the means of a conditional association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null) causes the rejection of the
hypothesis of conditional independence. Figure \ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status. The Chi-squared
test of independence rejects the null hypothesis: possibly, because
too many men who had premarital sex also had extramarital sex?
Finally, Figure \ref{fig:titanic} visualizes the `Survival on the Titanic' 
data using a double-decker plot. Here, a binary response (survival to the disaster)
shall be explained by other factors (class, gender, and age). The blue
boxes represent the proportion of survived passengers in a particular
stratum. The proportions of saved women and children are indeed higher
than those of men, but they clearly decrease from the 1st to the 3rd class. In
addition, the proportion of saved men in the 1st class is higher than
in the others.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(UCBAdmissions, panel = cotab_assoc, shade = TRUE, legend = FALSE)
@
\caption{Conditional association plot for the \data{UCBAdmissions} data.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

@@

This paper describes the strucplot framework provided by the \pkg{vcd} package
for the \proglang{R} system and environment for statistical computing
and graphics, available from the Comprehensive \proglang{R} Archive Network (\url{http://cran.r-project.org/}).
The framework integrates displays such as mosaic, association, and
sieve plots by their unifying property of being 
flat representations of contingency tables.
These basic plots, as well as specialized displays for conditional
independence, can be used both for exploratory visualization and
model-based analysis. Exploratory techniques include specialized displays for
the bivariate case, as well as pairs and trellis plot-like displays for
higher-dimensional tables. Model-based tools include 
methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (logistic regression,
log-linear models). Additionally, two of the frameworks' further strengths are its
flexibility and extensibility: graphical appearance aspects such as shading,
labeling, and spacing are modularized by means of `graphical
appearance functions', allowing fine-granular customization and
user-level extensions.

The remainder of the paper is organized as follows. In
Section \ref{sec:strucplot}, we give an overview of the strucplot
framework, describing the hierarchy of the main components and the
basic functionality. 
In Section \ref{sec:shading}, we demonstrate how
(residual-based) shadings support the visualization of log-linear models and
the results of independence tests. Also, we explain step-by-step how
the concepts of generating functions and graphical appearance control
functions can be combined to provide a flexible customization of
complex graphical displays as created by the strucplot framework.
Sections \ref{sec:labeling} and \ref{sec:spacing} discuss in detail the labeling
and spacing features, respectively, before Section \ref{sec:conclusion}
concludes the work.

@@

The strucplot framework is highly modularized: Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot. These are examples of `graphical appearance control'
(`grapcon') functions. They are created by generating functions
(`grapcon generators'), allowing
flexible parameterization and extensibility (Figure \ref{fig:struc}
only shows the generators). The generator names
follow the naming convention \code{\emph{group\_foo}()},  
where (\code{\emph{group}}) reflects the group the
generators belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\emph{foo}()}, 
\code{labeling\_\emph{foo}()}, and \code{legend\_\emph{foo}()}) 
directly produce graphical output (i.e., ``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\emph{foo}()} and \code{shading\_\emph{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\emph{foo}()} implement the core functionality, creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulae. Finally, on the fourth
level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

@@

As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous variable (gender). The
`flattened' contingency table can be obtained using the
\codefun{structable} function (quite similar to \codefun{ftable} in
base \proglang{R}, but allowing the specification of split directions):

@@

Let us first visualize the contingency table by means of a mosaic plot.
% \citep{vcd:Hartigan+Kleiner:1984} which is basically 
% an area-proportional visualization of (typically, observed) frequencies, composed
% of tiles (corresponding to the cells) created by recursive
% vertical and horizontal splits of a square. Thus, the area of each tile
% is proportional to the corresponding cell entry \emph{given} the
% dimensions of previous splits. 
Figure \ref{fig:observed} depicts the effect of

@@

%\setkeys{Gin}{width=0.75\textwidth}
\begin{figure}[p]
\begin{center}
<<Observedfig,fig=TRUE,echo=FALSE,height=6>>=
<<Observed>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data.}
\label{fig:observed}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<Expectedfig,fig=TRUE,echo=FALSE,height=6>>=
<<Expected>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}
%\setkeys{Gin}{width=0.7\textwidth}

@@

\noindent (Note that \code{hec}, a \class{structable} object, already includes a
splitting information which simply gets overloaded in this example.) 
For compatibility with \codefun{mosaicplot} in base \proglang{R},
the \codefun{mosaic} function also allows the use of a \code{"direction"} 
argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figure \ref{fig:split}):

@@

So far, we have visualized full tables. For objects of class \class{table},
conditioning on levels (i.e., choosing a table subset for fixed levels
of the conditioning variable(s)) is simply done by indexing.
However, subsetting \class{structable} objects is more restrictive 
because of their inherent conditional structure. Since the variables
on both the row and the columns side are nested, 
conditioning is only possible ``outside-in'':

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=12>>=
<<Variables>>
@
\caption{Two mosaic displays put side-by-side, visualizing the 
distribution of hair and eye color, given gender. The marginal
distribution of `Sex' cannot be seen.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
<<Conditioning>>
@
\caption{Mosaic plot again visualizing the 
distribution of hair and eye color, given gender, this time using a
single mosaic plot. In contrast to Figure \ref{fig:parttable}, this
plot also visualizes the marginal distribution of `Sex'.}
\label{fig:conditioning}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
<<cotabplot>>
@
\caption{Conditional table plot for the \data{HairEyeColor} data,
  again visualizing the distribution of hair and eye color, given
  gender, using separate mosaic displays like the `manual' plot 
  in Figure \ref{fig:parttable}.}
\label{fig:cotabplot}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

This flexible way of specifying graphical parameters is the basis for a suite of
shading functions that modify the tiles' appearance with respect to a vector of 
residuals, resulting from deviations of observed from expected
frequencies under a given log-linear model. 
The idea is to visualize at least sign and absolute size of the residuals, but some
shadings, additionally, indicate overall significance. One particular
shading, the maximum shading, even allows to identify those cells that
cause the rejection of the null hypothesis.

@@

\begin{figure}[p]
\begin{center}
<<continuousfig,fig=TRUE,echo=FALSE,height=7>>=
<<continuous2>>
@
\caption{The \data{HairEyeColor} data with continuous shading.}
\label{fig:continuous}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<bundesligafig,fig=TRUE,echo=FALSE,height=7>>=
<<bundesliga>>
@
\caption{Non-significant $\chi^2$ test using part of the \data{Bundesliga} data.}
\label{fig:bundesliga}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
<<friendlyfig,fig=TRUE,echo=FALSE,height=7>>=
<<friendly>>
@
\caption{The \data{Bundesliga} data for 1995 using the Friendly
  shading and a legend with fixed bins.}
\label{fig:friendly}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<hclshading1fig,fig=TRUE,echo=FALSE,height=6>>=
<<hclshading1>>
@
\caption{The \data{HairEyeColor} data, using default HCL color palette.}
\label{fig:hclshading1}
\end{center}
\end{figure}

\begin{figure}[p]
\begin{center}
<<hclshading2fig,fig=TRUE,echo=FALSE,height=6>>=
<<hclshading2>>
@
\caption{The \data{HairEyeColor} data, using a custom HCL color palette.}
\label{fig:hclshading2}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<rotatefig,fig=TRUE,echo=FALSE,height=6>>=
<<rotate>>
@
\caption{Rotating labels.}
\label{fig:rotating}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<repeatfig,fig=TRUE,echo=FALSE,height=6>>=
<<repeat>>
@
\caption{Inhibiting the repetition of levels.}
\label{fig:repeat}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<marginsfig,fig=TRUE,echo=FALSE,height=6>>=
<<margins>>
@
\caption{Changes in the margins.}
\label{fig:margins}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<boxesfig,fig=TRUE,echo=FALSE,height=6>>=
<<boxes>>
@
\caption{Boxes and Clipping.}
\label{fig:boxes}
\end{center}
\end{figure}

@@

Spacing of strucplot displays is customizable in a similar way than shading. The
\code{spacing} argument of the \codefun{strucplot} function takes a
list of \class{unit} vectors, one for each dimension, specifying the
space between the tiles corresponding to the levels. As an example,
consider the introductory example of the \data{Arthritis} data 
(Figure \ref{fig:arthritis}). Since we are interested in the effect of
the medicament in the placebo and treatment groups, a mosaic plot is
certainly appropriate to visualize the three levels of `Improved' in
the two `Treatment' strata. Another conceptual approach is to use
spine plots with highlighting \citep{vcd:hummel:1996}. A spine plot is
a variation of a bar plot where the heights of the bars are held
constant, whereas the widths are used to represent the number of cases
in each category. This is equivalent to a mosaic plot for a one-way table.
If a second (indicator) variable is highlighted in a spine plot, we
obtain a display equivalent to a simple mosaic display for a two-way
table, except that no space between the levels of the highlighted
variable is used.
In the \data{Arthritis} example, we will highlight patients with `Marked' 
improvement in both groups. To obtain such a display within 
the strucplot framework\footnote{The \pkg{vcd} package also provides 
a specialized \codefun{spine} function for spine plots and spinograms.}, it
suffices to set the space between the `Improved' tiles to 0 (see
Figure \ref{fig:artspine}):

<<artspine,echo=TRUE,results=verbatim>>=
(art <- structable(~Treatment + Improved, data = Arthritis, split_vertical = TRUE))
(my_spacing <- list(unit(0.5, "lines"), unit(c(0, 0), "lines")))
my_colors <- gpar(fill = c("lightgray", "lightgray", "black"))
mosaic(art, spacing = my_spacing, gp = my_colors)
@

@@

\begin{appendix}

\section{Data sets}
\label{sex:data}
